\input{../assignmentHeader.tex''}

\title{Economics 628 : Assignment 2}
\author{Paul Schrimpf}
\date{Due: September 26th, 2018}

\begin{document}
\maketitle


\begin{problem}$\;$
  In this problem, you'll reproduce and extend the analysis of
  \cite{qsw2018}. Begin by reading the paper.
  \begin{enumerate}
  \item What is the exogeneity assumption in equation (1) of section
    IV.B? How does the use of controls affect the plausibility of this
    assumption?
  \item Reproduce the results in Table 2. The authors' data and Stata
    code is on the AER website.
    \href{https://github.com/ubcecon/ECON628_2018/tree/master/paul/assignments/02/qsw-dataPrep.R}
    {qsw-dataPrep.R} 
    downloads and prepares the data for analysis.\footnote{This is
      mostly a direct translation of the authors' Stata code into
      R. As such, it's not written in an idiomatic R style.}
    \footnote{If you
      run on this on a non-Linux system you will have to change lines
      13-15 or just manually rename the ``data'' folder to ``qsw2018.''}
    \href{https://github.com/ubcecon/ECON628_2018/tree/master/paul/assignments/02/qsw-table2.R}
    {qsw-table2.R} estimates the model in column 3 of table 2.
  \item Next, we will use Lasso to allow for a more flexible
    specification of controls in equation (1). However, with panel
    data and a model with fixed effects, the Lasso must be modified
    slightly. Explain how the cluster-Lasso estimator of
    \cite{bchk2016} differs from the Lasso estimator for independent
    observations referenced in the notes. Alternatively, write code
    that calculates the cluster Lasso estimate. 
  \item
    \href{https://github.com/ubcecon/ECON628_2018/tree/master/paul/assignments/02/rlassoPanel.R}
    {rlassoEffectsPanel.R}
    implements the cluster Lasso estimator with fixed effects. Use it
    to estimate equation (1) while including many 
    more controls. Discuss the results. Did using Lasso add anything
    to this example?
  \end{enumerate}
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\bibliographystyle{jpe}
\bibliography{/home/paul/ECON628_2018/paul/01-machineLearningAndCausalInference/ml.bib}

\end{document}
